Lake Photic Zone Temperature across the Conterminous United States

### 2020-06-27 manuscript/manuscript.Rmd
merged Betty's edits, and ignored manuscript cache.

---
title: "Lake Photic Zone Temperature across the Conterminous United States"
author:
- Kreakie, B. J. ^\*^ ^*1*^, Milstead, W. Bryan. ^*1*^, Hollister. J. W. ^*1*^, Shivers,
  S. ^*2*^, 
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: kreakie.betty@epa.gov*'

@@

```{r model, include=FALSE, eval=TRUE, cache = TRUE}
source(here::here("R/model.R"), encoding = "UTF-8")
```
```{r figures, include=FALSE}
source(here::here("R/figures.R"), encoding = "UTF-8")
```

@@

As a proxy for directly measured ambient air temperature, we used the PRISM AN81d dataset (PRISM Climate Group, Oregon State University, http://prism.oregonstate.edu, created 07 Nov 2018).  This dataset provides interpolated daily temperature estimates (mean, maximum, and minimum) for 4km grids in the conterminous United States for 1981 to the present (see: http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf).  PRISM takes advantage of measured climate variables to interpolate point data to spatially defined grids using regression techniques and expert knowledge (Daly et al 2008).  For our study, we used the “prism” R package (Hart & Bell) to download the mean daily temperatures for the PRISM grid cells corresponding to the centroids of all NLA lakes included in this study for dates between 01 May and 30 September of 2007 and 2012.

@@

Using the 2007 and 2012 NLA data, we have successfully built a simple yet robust model of lake surface temperature for the conterminous United States. The final model has a mean square error of 2.19 and an adjusted R^2 of 0.88. The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy. Given the importance of temperature to a lake ecosystem, especially cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers.  Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be straightforwardly estimated based on basic ambient temperature and location information.

@@

For the final model, average ambient air temperature and the average temperature of the prior 30 days were selected.  Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model.  It is likely that the three and seven day averages are not providing the model with unique information.  Typical the three and seven day averages are fairly close the day-of average.  Ambient air temperature does not change drastically over such a short time scale.  Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model.  Yet the longer term 30-day average does have an impact on lake photic zone temperature.  The 30-day average provides us with this information about the temperature intensity leading up to the sample date.  Not just how far along into summer; that information would be captured in the date parameter.  But more specifically, this measures the long term thermal heating happening at a site.

In addition to the several derived air temperature variables, we included land-use/land cover variables in our initial variable selection process.  Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development (need to look at Jeff’s package to see how this is measured).  This variables were included based on the hypothesis that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes.  Yet neither of these variables were selected in the final model.  Even though the land-use variables were not selected that does not mean development and impervious are not impactful.  This urban-heat effect on lakes may have been adequately captured in the average ambient air temperature.  Therefore, making the land-use variables redundant.  Regardless these variables did not independently contribute to the model’s accuracy.

Insert paragraph on model shortcomings

>>>>

---
title: "Lake Photic Zone Temperature across the Conterminous United States"
author:
- Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead, W. Bryan. ^*1*^,
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: kreakie.betty@epa.gov*'

@@

```{r model, include=FALSE, eval=TRUE, cache=TRUE}
source(here::here("R/model.R"), encoding = "UTF-8")
```
```{r figures, include=FALSE, cache=TRUE}
source(here::here("R/figures.R"), encoding = "UTF-8")
```

@@

We included numerous predictor variables that are hypothesized to impact lake photic zone. As a proxy for directly measured ambient air temperature, we used the PRISM AN81d dataset (PRISM Climate Group, Oregon State University, http://prism.oregonstate.edu, created 07 Nov 2018).  This dataset provides interpolated daily temperature estimates (mean, maximum, and minimum) for 4km grids in the conterminous United States for 1981 to the present (see: http://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf).  PRISM takes advantage of measured climate variables to interpolate point data to spatially defined grids using regression techniques and expert knowledge (Daly et al 2008).  For our study, we used the “prism” R package (Hart & Bell) to download the mean daily temperatures for the PRISM grid cells corresponding to the centroids of all NLA lakes included in this study for dates between 01 May and 30 September of 2007 and 2012.

@@

Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States. The final model has a mean square error of 2.19 and an adjusted R^2 of 0.88. The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy. Given the importance of temperature to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers. Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be straightforwardly estimated based on basic ambient temperature and location information.

@@

The final model included, average ambient air temperature and the average temperature of the prior 30 days. Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model. It is likely that the three and seven day averages are not providing the model with unique information. Typical the three and seven day averages are fairly close the day-of average. Ambient air temperature does not change drastically over such a short time scale. Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model. Yet the longer term 30-day average does have an impact on lake photic zone temperature. The 30-day average provides us with this information about the temperature intensity leading up to the sample date. Not just how far along into summer; that information would be captured in the date parameter. But more specifically, this measures the long term thermal heating happening at a site.

In addition to the several derived air temperature variables, we included land-use/land cover variables in our initial variable selection process. Specifically, we calculated the percent impervious surface for a 3km lake buffer and a measure of shoreline development. These variables were included based on the hypothesis that higher amounts of development and therefore impervious surface surrounding a lake would lead to higher temperatures in lakes. Yet neither of these variables were selected in the final model. Even though the land-use variables were not selected that does not mean development and impervious are not impactful. This urban-heat effect on lakes may have been adequately captured in the average ambient air temperature. Therefore, making the land-use variables redundant. Regardless these variables did not independently contribute to the model’s accuracy.
